Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2018-10-04 paper/paper.tex
A few small changes up through about half of results. One or two sentences I think add a little bit of clarity. Also changed wildtype to wild-type (adjective) or wild type (noun)

\begin{abstract}
The outcome of viral infection is extremely heterogeneous at the cellular level, and infected cells only sometimes activate innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
\end{abstract}

@@

Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wildtype,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection, since they do not determine full sequences of the virions that infect individual cells.
Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wildtype'' or ``mutant'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain the transcriptomes and full sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a nominally ``wildtype'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wildtype virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wildtype virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich for rare IFN+ positive cells}
A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the single-cell level~\citep{killip2017single}.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
This paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.

This rareness of IFN induction is not merely a feature of A549 cells in culture.
Prior studies using reporters have found that IFN expression is rare at the single-cell level in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes (ISGs), which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
But just a handful of cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.

@@

\figsupp[Few cells express detectable IFN in influenza-infected mice.]
{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wildtype C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one transcript detected).
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
\label{figsupp:mice}

@@

\figsupp[Expression of type I and type III IFN is highly correlated in A549 cells.]
{An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
The cells were then infected with two different stocks of ``wildtype'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
After 13 hours, cells were analyzed by flow cytometry.
As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
{\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}}
\label{figsupp:type_I_vs_III}

@@

To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B).
These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
Cells that activate the IFN reporter can enriched by magnetic-activated cell sorting (MACS) for the cell-surface protein, or identified by flow cytometry for the fluorescent protein.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics described below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wildtype and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes using the 10X Chromium platform.
{\bf (C)}
The entire cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

\figsupp[MACS enrichment of IFN+ cells.]
{
Example MACS enrichments of IFN+ influenza-infected cells.
A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wildtype WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
{\bf (A)}
After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
{\bf (B)}
Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
The qPCR was performed in quadruplicate (hence the four points for each sample).
}
{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
\label{figsupp:MACS}

\figdata{Genbank files giving sequences of the wildtype and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
\label{figdata:virus_seqs}

@@

We therefore developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
First, we generated a stock of virus that consisted of a mix of wildtype WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).

We next used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of the cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter.
Based on pilot experiments, we expected this MACS to enrich IFN+ cells to $\sim$5\% to 30\% of the total (\FIGSUPP[workflow]{MACS}).
To ensure the presence of IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA among the human cells called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes among infected cells.
{\bf (F)} Number of cells infected with wildtype virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
}
\label{fig:transcriptomics}

@@

\figsupp[Expression of ISGs in single infected and uninfected cells.]
{
For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
Each point represents one cell, and the Pearson correlation coefficient is shown.
IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression more often due to paracrine signaling that does not induce expression of IFN itself.
Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}}
\label{figsupp:ISG}

@@

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wildtype and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.

@@

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wildtype.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wildtype.
These results show that the viral defects in single IFN+ cells often cause increased IFN production.

>>>>

\begin{abstract}
The outcome of viral infection is extremely heterogeneous at the level of individual cells, including sparse activation of innate immunity.  
Here we assess how the genetic variation inherent in viral populations contributes to this heterogeneity.
We do this by developing a new approach to determine both the transcriptome and full-length sequences of all viral genes in single influenza-infected cells.
Infections that activate an innate-immune response in single cells are associated with viral defects that include amino-acid mutations, internal deletions, and failure to express key genes.  
However, immune activation remains stochastic in cells infected by virions with these defects, and sometimes occurs even in cells infected by virions that express unmutated copies of all genes.
Our work shows that the genetic variation present in influenza virus populations substantially contributes to but does not fully explain the heterogeneity in infection outcome and immune activation in single infected cells.
\end{abstract}

@@

Unfortunately for the host, influenza infection only rarely triggers IFN production by individual infected cells early in infection~\citep{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes influenza infection: infected cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\citep{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral replication and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially as real influenza infections are typically initiated by just a few virions entering a few cells~\citep{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon homogeneous treatment with synthetic innate-immune ligands~\citep{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\citep{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, even if the \emph{consensus} sequence of a viral population is ``wild-type,'' individual virions will often have defects~\citep{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\citep{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\citep{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection, since they do not determine full sequences of the virions that infect individual cells.
Flow cytometry and fluorescent reporters only measure protein levels~\citep{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\citep{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation, despite the fact that such mutations are common regardless of whether the cells are infected with a nominally ``wild-type'' or ``mutant'' stock of virus.

Here we develop a new approach to measure both the full transcriptome and the sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based single-cell transcriptomics and full-length PacBio sequencing of viral genes.
We obtain the transcriptomes and full sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Even though we used a nominally ``wild-type'' stock of virus, two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by truly wild-type virions having a tighter distribution of viral transcriptional burden.
Viral defects are especially common in IFN+ cells, and we identify at least four defects that cause increased IFN induction: amino-acid substitutions in the viral polymerase (PB1), absence of the NS gene, amino-acid substitutions in NS1, and internal deletions in polymerase genes.
However, viral genetic variation does not fully explain the cell-to-cell heterogeneity, as even truly wild-type virions sometimes induce IFN, and no viral mutations deterministically induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich for rare IFN+ positive cells}
A challenge in studying IFN induction by influenza virus is that infection only rarely triggers expression of anti-viral IFN~\citep[type I and type III;][]{kotenko2017contribution} at the level of single-cells ~\citep{killip2017single}.
For instance, we previously performed single-cell transcriptomics on hundreds of influenza-infected A549 human lung epithelial cells at early times (6 to 10 hours) post-infection~\citep{russell2018extreme}.
Only 1 of 368 infected cells expressed detectable type I or type III IFN (\FIG{IFNrare}A).
This paucity of IFN+ cells was not due simply to a poor detection threshold, since $>$100 IFN transcripts were detected in the one IFN+ cell~\citep{russell2018extreme}.

This rareness of IFN induction is not merely a feature of A549 cells in culture.
Prior studies using interferon reporter constructs have found that IFN expression is rare in influenza-infected mice at 24 and even 48 hours post-infection~\citep{kallfass2013visualizing}.
In support of these findings, we re-analyzed the single-cell transcriptomics of influenza-infected mice by \citet{steuerman2018dissection}.
As shown in \FIGSUPP[IFNrare]{mice}, many infected cells in the mice expressed IFN-stimulated genes (ISGs), which are induced by autocrine and paracrine signaling~\citep{stetson2006type,honda2006type}.
But just a handful of cells detectably expressed IFN itself, which is only induced by direct cellular detection of viral infection~\citep{stetson2006type,honda2006type}.

@@

\figsupp[Few cells express detectable IFN in influenza-infected mice.]
{A re-analysis of the data from \citet{steuerman2018dissection}'s single-cell transcriptomics of cells from influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \citet{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \citet{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \citet{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in \citet{russell2018extreme} and the current study, respectively).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, and only at low levels (one transcript detected).
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
{\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}}
\label{figsupp:mice}

@@

\figsupp[Expression of type I and type III IFN is highly correlated in A549 cells.]
{An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
After 13 hours, cells were analyzed by flow cytometry.
As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.}
{\includegraphics[width=0.5\textwidth]{figures/IFN_stochastic/IFN_reporter/IFNbeta_IFN_lambda_correlated.pdf}}
\label{figsupp:type_I_vs_III}

@@

To efficiently identify and enrich rare IFN+ cells, we integrated IFN reporters into the genome of A549 cells (\FIG{IFNrare}B) using lentiviral transduction and subsequently propagated these reporters from single clones.
These reporters consisted of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein~\citep[LNGFR$\Delta$C;][]{bonini1997hsv,ruggieri1997cell} followed by a fluorescent protein.
Cells that activate the IFN reporter can enriched by magnetic-activated cell sorting (MACS) for the cell-surface protein, or identified by flow cytometry for the fluorescent protein.
The reporters were efficiently activated by a strain of Sendai virus~\citep{strahle2006sendai} that potently induces IFN (\FIGSUPP[IFNrare]{reporter_validation}), and activation of the type I and type III IFN reporters was highly correlated in single influenza-infected A549 cells (\FIGSUPP[IFNrare]{type_I_vs_III}; further validated by the single-cell transcriptomics described below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

@@

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\FIGSUPP[workflow]{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes using the 10X Chromium platform.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio.
{\bf (E)}wild-type
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

\figsupp[MACS enrichment of IFN+ cells.]
{
Example MACS enrichments of IFN+ influenza-infected cells.
A549 cells with the \textit{IFNB1} LNGFR$\Delta$C-mNeonGreen reporter were infected with wild-type WSN influenza (two different viral stocks) at a target MOI of 0.1 TCID50 per cell.
After infection had proceeded for 12 hours, the cells were twice magnetically sorted for LNGFR$\Delta$C expression over magnetic columns as detailed in the methods for the single-cell sequencing experiment.
{\bf (A)}
After sorting, the populations were analyzed by flow cytometry for IFN expression using the mNeonGreen fluorescent protein.
The plots show the distribution of fluorescence in the original population, the flow-through from the first column, and the MACS-sorted positive population after two columns.
As indicated by the percentages shown for the original and MACS-sorted population, this process led to substantial enrichment in IFN+ cells.
We expect that the IFN sorting for the actual single-cell sequencing led to similar enrichment, although we could not directly quantify this as the sorted cells in that case were immediately used for the sequencing and so could not be analyzed by flow cytometry.
{\bf (B)}
Analysis of expression of IFNB1 (relative to the housekeeping gene L32) by qPCR in the positive (IFN enriched) and negative (IFN depleted) populations from panel (A).
The qPCR validates a roughly 50- to 100-fold enrichment in total IFNB1 expression.
The qPCR was performed in quadruplicate (hence the four points for each sample).
}
{\includegraphics[width=\textwidth]{figures/MACS/MACS.pdf}}
\label{figsupp:MACS}

\figdata{Genbank files giving sequences of the wild-type and synonymously barcoded viruses are in \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn.gb} and \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/blob/master/data/flu_sequences/flu-wsn-double-syn.gb}.}
\label{figdata:virus_seqs}

@@

We therefore developed the approach in \FIG{workflow} to determine the entire transcriptome \emph{and} the full sequences of the infecting virions in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\FIGDATA[workflow]{virus_seqs}).
These viral barcodes allow us to identify co-infection of individual cells from single-cell transcriptomic data~\citep{russell2018extreme}, and also provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).

We next used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of the cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that had activated the IFN reporter.
Based on pilot experiments, we expected this MACS to enrich IFN+ cells to $\sim$5\% to 30\% of the total (\FIGSUPP[workflow]{MACS}).
To ensure the presence of infected IFN- cells after MACS enrichment, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\citep{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell, and the blue rug plot at the left of each panel shows the distribution of viral mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\FIGSUPP[transcriptomics]{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes among infected cells.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\FIGSUPP[transcriptomics]{ISG}).
}
\label{fig:transcriptomics}

@@

\figsupp[Expression of ISGs in single infected and uninfected cells.]
{
For each cell, we quantified ISG expression as the total fraction of cellular mRNAs derived from four prototypical ISGs (IFIT1, ISG15, CCL5, and Mx1). 
{\bf (A)} The histograms show the distribution of ISG expression taken across infected (top) and uninfected (bottom) cells.
We heuristically classify as ISG+ cells with $>10^{-3}$ of their cellular mRNA from ISGs, and color these cells red.
Comparison to \FIG{transcriptomics}G shows that substantially more cells are ISG+ than IFN+, both among infected and uninfected cells.
This is probably because paracrine signaling can induce ISG expression in cells that are not themselves expressing IFN~\citep{stetson2006type,honda2006type}.
{\bf (B)} Correlation between the fraction of cellular mRNA derived from IFN and ISGs.
Each point represents one cell, and the Pearson correlation coefficient is shown.
IFN and ISG expression are more correlated for infected than uninfected cells, probably because in the latter the ISG expression is more often due to paracrine signaling that does not induce expression of IFN itself.
Among both the infected and uninfected populations, there are many cells with high expression of ISGs and little expression of IFN, but no cells that express high levels of IFN without also substantially expressing ISGs.
}
{\includegraphics[width=0.8\textwidth]{figures/single_cell_figures/p_isg.pdf}}
\label{figsupp:ISG}

@@

By examining the synonymous viral barcodes near the 3' termini of transcripts, we determined that 38\% of cells were co-infected with wild-type and synonymously barcoded virions (\FIG{transcriptomics}F; cells called as co-infected if a binomial test rejected null hypothesis that $\ge$95\% of viral mRNA is from one viral barcode variant).
From \FIG{transcriptomics}F, we estimate~\citep{bloom2018estimating} that 63\% of infected cells are co-infected, implying that 25\% are co-infected with two virions with the same viral barcode (such co-infections cannot be identified from the transcriptomic data).
Interestingly, this co-infection rate is higher than expected from the relative numbers of infected and uninfected cells (\FIG{transcriptomics}C) if infection is a Poisson process.
This discrepancy could arise if the MACS for IFN+ cells also enriches for co-infected cells, if infection itself is not truly Poisson, or if co-infection increases the likelihood that we identify a cell as infected given the thresholds in \FIG{transcriptomics}C.
This high rate of observed coinfection can also partially explain the relative completeness of many viral infections in our dataset, as absence of a segment from one viral particle can be readily complemented by a second, infecting, particle~\citep{russell2018extreme}.

@@

We tested the rate of IFN induction by each viral stock using the reporter cells.
\FIG{validation} shows that five of the eight mutant viral stocks induced IFN more frequently than wild type.
The strongest IFN induction was by the NS1stop virus, but the PB1 internal deletion and three of the point-mutant viruses (PB1-D27N, PB1-T677A, and NS1-A122V) also induced significantly more IFN than wild-type.
These results show that the viral defects in single IFN+ cells often cause increased IFN production.
